\texttt{batman}: BAsic Transit Model cAlculatioN in Python

### 2015-07-23 manuscript/ms.tex
update text and fig3

\title{BATMAN}

@@

\begin{abstract}
I introduce \texttt{batman}, a Python package for modeling exoplanet transit light curves.  \texttt{batman} supports calculation of light curves for any radially symmetric stellar limb darkening law, using a new integration algorithm for models that cannot be calculated analytically.  The code uses C extension modules to speed up model calculation and is parallelized with OpenMP.  (FIXME: say how long it takes to compute a million models). Model light curves can be computed with accuracy better than 0.0001 parts per million, ensuring that the community is prepared for the extraordinarily precise transit light curves we anticipate measuring with upcoming facilities.
\end{abstract}

\keywords{}

\section{Introduction}
When an extrasolar planet transits in front of its host star, a small fraction of the stellar flux is blocked by the planet.  Precise measurements of stellar brightness as a function of time (light curves) can therefore reveal transiting planets. Over a thousand planets have been discovered with this technique \citep[e.g.][]{rowe2015}.  Transit light curves can be used to characterize the planets in detail, enabling measurements of planetary radius, orbital elements, and even atmospheric temperature structure and composition \citep{seager00, seager03}.  %This characterization depends on a foundation of accurate transit light curve models.

Transit light curve models are a very basic part of the toolkit for exoplanet characterization, but they are not trivial to compute.  The model must account for the planet's size and position on the sky, as well as stellar limb darkening, which causes the apparent brightness of the stellar disk to decrease from center to edge.  The  stellar intensity profile can be fit with several functional forms, including a linear limb darkening law \citep{schwarzschild06}, quadratic \citep{kopal50}, square-root \citep{diaz-cordoves92}, logarithmic \citep{klinglesmith70}, exponential \citep{claret03}, and four-parameter nonlinear \citep{claret00}.  For some of these profiles, model transit light curves can be calculated analytically \citep{mandel02, gimenez06, abubekerov13}.  Other profiles do not have analytic solutions, and models must be calculated by numeric integration of the stellar intensity over the disk of the planet.

@@

In this paper, I introduce the Python package \texttt{batman}.  This package enables fast computation of transit light curves for any radially symmetric limb darkening law, and currently supports uniform, linear, quadratic, logarithmic, exponential, and nonlinear limb darkening. Light curves for the first three of these are calculated analytically based on the formalism from \cite{mandel02}.  Models for the remaining cases are computed with an efficient new integration scheme, described in \S\,\ref{section:algorithm}. I discuss \texttt{batman}'s features and performance in \S\,\ref{section:batman} and conclude in \S\,\ref{section:summary}.

%This package implements an efficient new algorithm (described in \S\,\ref{section:algorithm}) to calculate transit light curve models for any radially symmetric limb darkening profile.  The transit models are computed with C extension modules written with the Python/C API, and the code is parallelized at the C level with OpenMP.  The package therefore   mostly widely used programming language

@@

See Figure\,\ref{fig:integration} for a schematic illustrating the geometry of the integration.  The advantage of this integration scheme is that the only error introduced is due to approximating the stellar intensity as constant over the differential area element $\Delta A = A(x_i, r_p, d) - A(x_{i-1}, r_p, d)$.  Computation of $\Delta A$ is expensive, but it needs to be calculated relatively few times ($<<10^{6}$ for sub-ppm accuracy). This makes the integration faster than one with a simpler area element (e.g., $\Delta A = \Delta x \Delta y$), that requires a much smaller step size to achieve the same accuracy.

@@

\section{The \texttt{batman} package}
\label{section:batman}
The Python package \texttt{batman} implements the algorithm described in \S\,\ref{section:algorithm} to calculate model transit light curves.  \texttt{batman} is an open source project and is being developed on GitHub.  Full documentation is available at \blue{\url{https://github.com/lkreidberg/batman}}. I summarize the main capabilities of the package here.

@@

The square-root, logarithmic, exponential, nonlinear, and custom models are computed with the numeric integration scheme from \S\,\ref{section:algorithm}.  The uniform, linear, and quadratic models are calculated analytically, with code based on the Fortran routines \texttt{occultquad.f} and \texttt{occultuniform.f} provided by \cite{mandel02}. For the analytic models, I follow \cite{eastman13} and use the algorithm from \cite{bulirsch65} to improve the calculation speed and accuracy for elliptic integrals of the third kind.

@@

I also tested the accuracy of the widely-used \cite{mandel02} code that uses Numerical Recipes functions to calculate elliptic integrals \citep{press92}.  The accuracy was better than 0.005 ppm for most input values; however, for the case $r_p - d < \epsilon$, the error in the light curve exceeded 2 ppm. By contast, the \cite{bulirsch65} algorithm for elliptic integrals is well-behaved for this case and also faster.

@@

The computationally intensive sections of the code (including all of the transit model calculation) are written as C extension modules with the Python/C API, which improves the performance by roughly two orders of magnitude over a pure Python implementation.  The code also includes the option to parallelize at the C level with OpenMP, which further speeds up the calculation. The number of processors is specified by the user.  \texttt{batman} will raise an exception if the user attempts to parallelize a calculation on a system where OpenMP is not supported.

I tested \texttt{batman}'s performance over a range of typical use cases using a 1.7 GHz Intel Core i5 processor.  In Figure\,\ref{fig:performance}, I show the execution time versus error tolerance for a single transit light curve calculation using a nonlinear intensity profile, compared to the execution time for a quadratic model computed analytically.  The test case consists of 100 points evenly sampled in time during the planet's transit. I used physical parameters for the transiting planet GJ\,1214b \citep{kreidberg14a}. The stellar intensity profile is the same for the nonlinear and quadratic models: the nonlinear limb darkening coefficients are $(0.0, 0.7, 0.0, -0.3)$ and the quadratic coefficients are $(0.1, 0.3)$.  For this case, reaching a truncation error of 1.0 ppm via integration is just 3 times slower than the analytic calculation. Decreasing the truncation error by a factor of 10 increases the computation time by a factor of three.

@@

The analytic solution is challenging to compute because it uses the Appell F1 hypergeometric function.  This function is only convergent for certain regions of parameter space and must be calculated with analytic continuation for other cases. \cite{colavecchia04} provide a Fortran library for computing Appell F1.  I used this library to implement the analytic model for nonlinear limb darkening. However, the returned Appell F1 values are not accurate for all input parameters (based on a comparison with \texttt{Mathematica} and the pure Python library \texttt{mpmath}).  Even for cases where F1 is correct, the computation is FIXME times slower than numeric integration (for an error tolerance of FIXME).  I concluded that integration is a faster and easier solution than analytic models for nonlinear limb darkening.

\section{Summary}
\label{section:summary}
I introduced a new algorithm for computing transit light curves for any radially symmetric stellar limb darkening law.  I also described the Python package \texttt{batman}, a fast and versatile code to generate transit light curve models.  \texttt{batman} can calculate light curves with accuracy better than 0.0001 ppm, ensuring that we are prepared for the extraordinarily precise data we anticipate from upcoming facilities. (FIXME don't repeat abstract)

\acknowledgments

>>>>

\title{\texttt{batman}: BAsic Transit Model cAlculatioN in Python}

@@

\begin{abstract}
I introduce \texttt{batman}, a Python package for modeling exoplanet transit light curves.  The \texttt{batman} package supports calculation of light curves for any radially symmetric stellar limb darkening law, using a new integration algorithm for models that cannot be calculated analytically.  The code uses C extension modules to speed up model calculation and is parallelized with OpenMP.  For a typical light curve with 100 data points in transit, \texttt{batman} can calculate one million quadratic limb-darkened models in three minutes with a single 1.7 GHz Intel Core i5 processor.  Models can be computed with accuracy better than 0.001 parts per million, ensuring that the community is prepared for the precise transit light curves we anticipate measuring with upcoming facilities.  The code is open source and publicly available at \texttt{\blue{\url{https://github.com/lkreidberg/batman}}}.
\end{abstract}

\keywords{methods: data analysis -- methods: numerical}

\section{Introduction}
Exoplanet transit light curves can be used to characterize planetary systems in detail, enabling measurements of the planet's radius, orbital elements, and even atmospheric temperature structure and composition \citep{seager00, seager03}.  A number of current and planned observational facilities -- including \emph{K2}, \emph{TESS}, \emph{CHEOPS}, \emph{JWST}, and \emph{PLATO} -- will measure precise transit light curves for thousands of exoplanets that will shed light on planetary formation, evolution, and habitability.

Transit light curve models are a fundamental part of the toolkit for exoplanet characterization, but they are not trivial to compute.  The model must account for the planet's size and position on the sky, as well as stellar limb darkening, which causes the apparent brightness of the stellar disk to decrease from center to edge.  The  stellar intensity profile can be fit with several functional forms, including a linear limb darkening law \citep{schwarzschild06}, quadratic \citep{kopal50}, square-root \citep{diaz-cordoves92}, logarithmic \citep{klinglesmith70}, exponential \citep{claret03}, and four-parameter nonlinear \citep{claret00}.  For some of these profiles, model transit light curves can be calculated analytically \citep{mandel02, gimenez06, abubekerov13}.  Other profiles do not have analytic solutions, and models must be calculated by numeric integration of the stellar intensity over the disk of the planet.

@@

In this paper, I introduce the open source Python package \texttt{batman}.  This package enables fast computation of transit light curves for any radially symmetric limb darkening law, and currently supports uniform, linear, quadratic, logarithmic, exponential, and nonlinear limb darkening. Light curves for the first three of these are calculated analytically based on the formalism from \cite{mandel02}.  Models for the remaining cases are computed with an efficient new integration scheme, described in \S\,\ref{section:algorithm}. I discuss \texttt{batman}'s features and performance in \S\,\ref{section:batman} and conclude in \S\,\ref{section:summary}.

@@

See Figure\,\ref{fig:integration} for a schematic illustrating the geometry of the integration.  The advantage of this integration scheme is that the only error introduced is due to approximating the stellar intensity as constant over the differential area element $\Delta A = A(x_i, r_p, d) - A(x_{i-1}, r_p, d)$.  Computation of $\Delta A$ is expensive, but it needs to be calculated relatively few times ($<<10^{6}$ for sub-ppm accuracy). This makes the integration faster than a scheme with a simpler area element (e.g., $\Delta A = \Delta x \Delta y$), that requires a much smaller step size to achieve the same accuracy.

@@

\section{The \texttt{batman} package}
\label{section:batman}
The Python package \texttt{batman} implements the algorithm described in \S\,\ref{section:algorithm} and several analytic models to calculate transit light curves.  \texttt{batman} is an open source project and is being developed on GitHub.  Full documentation is available at \blue{\url{https://github.com/lkreidberg/batman}}. I summarize the main capabilities of the package here.

@@

The square-root, logarithmic, exponential, nonlinear, and custom models are computed with the numeric integration scheme from \S\,\ref{section:algorithm}.  The uniform, linear, and quadratic models are calculated analytically, with code based on the Fortran routines \texttt{occultquad.f} and \texttt{occultuniform.f} provided by \cite{mandel02}. For the analytic models, I follow \cite{eastman13} and use the algorithm from \cite{bulirsch65} to improve calculation speed and accuracy for elliptic integrals of the third kind.

@@

I also tested the accuracy of the widely-used \cite{mandel02} code that uses Numerical Recipes functions to calculate elliptic integrals \citep{press92}.  The accuracy was better than 0.005 ppm for most input values; however, for the case $r_p - d < \epsilon$, the error in the light curve exceeded 2 ppm. By contrast, the \cite{bulirsch65} algorithm for elliptic integrals is well-behaved for this case and also faster.

@@

Computationally intensive sections of code (including all of the transit model calculation) are written as C extension modules with the Python/C API, which improves the performance by roughly two orders of magnitude over a pure Python implementation.  \texttt{batman} also includes the option to parallelize at the C level with OpenMP, which further speeds up the calculation. The number of processors is specified by the user.  \texttt{batman} will raise an exception if the user attempts to parallelize a calculation on a system where OpenMP is not supported.

I tested \texttt{batman}'s performance over a range of typical use cases with a 1.7 GHz Intel Core i5 processor.  In Figure\,\ref{fig:performance}, I show the execution time versus error tolerance for a single transit light curve calculation using a nonlinear intensity profile, compared to the execution time for a quadratic model computed analytically.  The test case consists of 100 points evenly sampled in time during the planet's transit. I used physical parameters for the transiting planet GJ\,1214b \citep{kreidberg14a}. The stellar intensity profile is the same for the nonlinear and quadratic models: the nonlinear limb darkening coefficients are $(0.0, 0.7, 0.0, -0.3)$ and the quadratic coefficients are $(0.1, 0.3)$.  For this case, reaching a truncation error of 1.0 ppm via integration is just 3 times slower than the analytic calculation. Decreasing the truncation error by a factor of 10 increases the computation time by a factor of three.

@@

The analytic solution is challenging to compute because it uses the Appell F1 hypergeometric function.  This function is only convergent for certain regions of parameter space and must be calculated with analytic continuation for other cases. \cite{colavecchia04} provide a Fortran library for computing Appell F1.  I used this library to implement the analytic model for nonlinear limb darkening. However, the returned Appell F1 values are not accurate for all input parameters (based on a comparison with \texttt{Mathematica} and the pure Python library \texttt{mpmath}).  Even for cases where F1 is correct, the computation is over an order of magnitude slower than numeric integration (for an error tolerance of 0.1 ppm).  I concluded that integration is a faster and easier solution than analytic models for nonlinear limb darkening.

\section{Summary}
\label{section:summary}
I introduced a new algorithm for computing transit light curves for any radially symmetric stellar limb darkening law.  I also described the open-source Python package \texttt{batman}, a versatile code to generate model light curves. Uniform, linear, quadratic, logarithmic, exponential, and four-parameter nonlinear limb darkening laws are currently supported.  \texttt{batman} uses C extension modules to compute light curves and is parallelized with OpenMP to optimize performance.  Light curves can be calculated with accuracy better than 0.001 ppm, ensuring that the community is prepared to model the extraordinarily precise data we anticipate from upcoming facilities. \texttt{batman} is available at \texttt{\blue{\url{https://github.com/lkreidberg/batman}}} and is also hosted on the Python Package Index under the name \texttt{batman-package}.

\acknowledgments
I thank Jacob Bean, Kevin Stevenson, Eric Agol, Ethan Kruse, Geert Jan Talens, Thomas Beatty, Brett Morris, and Karl Fogel for their support in developing \texttt{batman}. I also thank contributors to SciPy, Matplotlib, and the Python Programming Language for software and services.  Support for this work was provided by a grant from the National Science Foundation through a Graduate Research Fellowship to L.K.
